How many speech sounds are there in the world's languages?

### 2023-06-21 paper.Rmd
updates

Aside from the acoustic-phonetic argument, evidence for infinite phonemes comes from studies on the cross-linguistic frequencies of speech sounds. Early evidence is found in a large cross-linguistic sample of phoneme inventories, where inventory is understood as the set of distinctive sounds attested in a language. This sample is described in the UCLA Phonological Segment Inventory Database (UPSID; @Maddieson1984, @MaddiesonPrecoda1990). @Maddieson1984 created a genealogically-balanced sample of the world's languages (*n* = 317) to investigate cross-linguistic properties of phonological systems and the relationship between aspects such as inventory size, contents, and structure. Maddieson noticed that certain segments are more or less likely to be present in inventories regardless of their size. That is, certain sounds are almost always present and others are almost never present. In other words, no segment occurs in every language and some very frequent segments occur in most languages. Moreover, nearly half of the reported segment types are singletons in the sample, i.e., one-off language-specific occurrences.^[Whether rare segments occur more or less frequently *within* a language is a language-by-language research question. See for example [@Everett2018].] While Maddieson did not explicitly test for it, his observations fit a well-attested pattern known as Zipf's law [@Zipf1936;@Zipf1949]. Zipf's law states that the frequency distribution of words in a language is inversely proportional to their frequency rank. The relationship is log-linear: high frequency words are few in number but account for most of the tokens in a text, and low frequency words are many, but occur rarely or even just once.^[Here is a plot showing Zipf's law in the first 10 million words in the Wikipedias from 30 different languages: https://en.wikipedia.org/wiki/Zipf%27s_law#/media/File:Zipf_30wiki_en_labels.png.] , Zipf's law states that the frequency of an element $f(r)$ is proportional to the inverse frequency rank of that element (modulated by an exponent $\alpha$) [@Piantadosi2014]:^[@Piantadosi2014 discusses in detail that word frequency distributions are not actually as simple as they seem according to Zipf's law. For example, semantics affects word frequency distributions. Moreover, observations of Zipf's law have been obscured by word frequency distribution plots that fail to address correlated errors. Zipfian distributions are not unique to language. They appear in various aspects of human society, such as music and aspects of the internet, as well as in other biological and physical systems. Strikingly, power law distributions can arise from nearly nothing [@Piantadosi2014, p. 1121].]

>>>>

Aside from the acoustic-phonetic argument, evidence for infinite phonemes comes from studies on the cross-linguistic frequencies of speech sounds. Early evidence is found in a large cross-linguistic sample of phoneme inventories, where inventory is understood as the set of distinctive sounds attested in a language. This sample is described in the UCLA Phonological Segment Inventory Database (UPSID; @Maddieson1984, @MaddiesonPrecoda1990). @Maddieson1984 created a genealogically-balanced sample of the world's languages (*n* = 317) to investigate cross-linguistic properties of phonological systems and the relationship between aspects such as inventory size, contents, and structure. Maddieson noticed that certain segments are more or less likely to be present in inventories regardless of their size. That is, certain sounds are almost always present and others are almost never present. In other words, no segment occurs in every language and some very frequent segments occur in most languages. Moreover, nearly half of the reported segment types are singletons in the sample, i.e., one-off language-specific occurrences.^[Whether rare segments occur more or less frequently *within* a language is a language-by-language research question. See for example @Everett2018.] While Maddieson did not explicitly test for it, his observations fit a well-attested pattern known as Zipf's law [@Zipf1936;@Zipf1949]. Zipf's law states that the frequency distribution of words in a language is inversely proportional to their frequency rank. The relationship is log-linear: high frequency words are few in number but account for most of the tokens in a text, and low frequency words are many, but occur rarely or even just once.^[Here is a plot showing Zipf's law in the first 10 million words in the Wikipedias from 30 different languages: https://en.wikipedia.org/wiki/Zipf%27s_law#/media/File:Zipf_30wiki_en_labels.png.] , Zipf's law states that the frequency of an element $f(r)$ is proportional to the inverse frequency rank of that element (modulated by an exponent $\alpha$) [@Piantadosi2014]:^[@Piantadosi2014 discusses in detail that word frequency distributions are not actually as simple as they seem according to Zipf's law. For example, semantics affects word frequency distributions. Moreover, observations of Zipf's law have been obscured by word frequency distribution plots that fail to address correlated errors. Zipfian distributions are not unique to language. They appear in various aspects of human society, such as music and aspects of the internet, as well as in other biological and physical systems. Strikingly, power law distributions can arise from nearly nothing [@Piantadosi2014, p. 1121].]

### 2023-06-21 paper.tex
adds paper
updates

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{06 June, 2023}

@@

Aside from the acoustic-phonetic argument, evidence for infinite
phonemes comes from studies on the cross-linguistic frequencies of
speech sounds. Early evidence is found in a large cross-linguistic
sample of phoneme inventories, where inventory is understood as the set
of distinctive sounds attested in a language. This sample is described
in the UCLA Phonological Segment Inventory Database (UPSID; Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}), Maddieson \& Precoda
(\protect\hyperlink{ref-MaddiesonPrecoda1990}{1990})). Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}) created a
genealogically-balanced sample of the world's languages (\emph{n} = 317)
to investigate cross-linguistic properties of phonological systems and
the relationship between aspects such as inventory size, contents, and
structure. Maddieson noticed that certain segments are more or less
likely to be present in inventories regardless of their size. That is,
certain sounds are almost always present and others are almost never
present. In other words, no segment occurs in every language and some
very frequent segments occur in most languages. Moreover, nearly half of
the reported segment types are singletons in the sample, i.e., one-off
language-specific occurrences.\footnote{Whether rare segments occur more
  or less frequently \emph{within} a language is a language-by-language
  research question. See for example
  (\protect\hyperlink{ref-Everett2018}{Everett 2018}).} While Maddieson
did not explicitly test for it, his observations fit a well-attested
pattern known as Zipf's law (\protect\hyperlink{ref-Zipf1936}{Zipf
1936}; \protect\hyperlink{ref-Zipf1949}{Zipf 1949}). Zipf's law states
that the frequency distribution of words in a language is inversely
proportional to their frequency rank. The relationship is log-linear:
high frequency words are few in number but account for most of the
tokens in a text, and low frequency words are many, but occur rarely or
even just once.\footnote{Here is a plot showing Zipf's law in the first
  10 million words in the Wikipedias from 30 different languages:
  \url{https://en.wikipedia.org/wiki/Zipf\%27s_law\#/media/File:Zipf_30wiki_en_labels.png}.}
, Zipf's law states that the frequency of an element \(f(r)\) is
proportional to the inverse frequency rank of that element (modulated by
an exponent \(\alpha\))
(\protect\hyperlink{ref-Piantadosi2014}{Piantadosi 2014}):\footnote{Piantadosi
  (\protect\hyperlink{ref-Piantadosi2014}{2014}) discusses in detail
  that word frequency distributions are not actually as simple as they
  seem according to Zipf's law. For example, semantics affects word
  frequency distributions. Moreover, observations of Zipf's law have
  been obscured by word frequency distribution plots that fail to
  address correlated errors. Zipfian distributions are not unique to
  language. They appear in various aspects of human society, such as
  music and aspects of the internet, as well as in other biological and
  physical systems. Strikingly, power law distributions can arise from
  nearly nothing (\protect\hyperlink{ref-Piantadosi2014}{Piantadosi
  2014: 1121}).}

>>>>

\title{How many speech sounds are there in the world's languages?}
\author{Steven Moran \& Nicholas A. Lester}
\date{21 June, 2023}

@@

Aside from the acoustic-phonetic argument, evidence for infinite
phonemes comes from studies on the cross-linguistic frequencies of
speech sounds. Early evidence is found in a large cross-linguistic
sample of phoneme inventories, where inventory is understood as the set
of distinctive sounds attested in a language. This sample is described
in the UCLA Phonological Segment Inventory Database (UPSID; Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}), Maddieson \& Precoda
(\protect\hyperlink{ref-MaddiesonPrecoda1990}{1990})). Maddieson
(\protect\hyperlink{ref-Maddieson1984}{1984}) created a
genealogically-balanced sample of the world's languages (\emph{n} = 317)
to investigate cross-linguistic properties of phonological systems and
the relationship between aspects such as inventory size, contents, and
structure. Maddieson noticed that certain segments are more or less
likely to be present in inventories regardless of their size. That is,
certain sounds are almost always present and others are almost never
present. In other words, no segment occurs in every language and some
very frequent segments occur in most languages. Moreover, nearly half of
the reported segment types are singletons in the sample, i.e., one-off
language-specific occurrences.\footnote{Whether rare segments occur more
  or less frequently \emph{within} a language is a language-by-language
  research question. See for example Everett
  (\protect\hyperlink{ref-Everett2018}{2018}).} While Maddieson did not
explicitly test for it, his observations fit a well-attested pattern
known as Zipf's law (\protect\hyperlink{ref-Zipf1936}{Zipf 1936};
\protect\hyperlink{ref-Zipf1949}{Zipf 1949}). Zipf's law states that the
frequency distribution of words in a language is inversely proportional
to their frequency rank. The relationship is log-linear: high frequency
words are few in number but account for most of the tokens in a text,
and low frequency words are many, but occur rarely or even just
once.\footnote{Here is a plot showing Zipf's law in the first 10 million
  words in the Wikipedias from 30 different languages:
  \url{https://en.wikipedia.org/wiki/Zipf\%27s_law\#/media/File:Zipf_30wiki_en_labels.png}.}
, Zipf's law states that the frequency of an element \(f(r)\) is
proportional to the inverse frequency rank of that element (modulated by
an exponent \(\alpha\))
(\protect\hyperlink{ref-Piantadosi2014}{Piantadosi 2014}):\footnote{Piantadosi
  (\protect\hyperlink{ref-Piantadosi2014}{2014}) discusses in detail
  that word frequency distributions are not actually as simple as they
  seem according to Zipf's law. For example, semantics affects word
  frequency distributions. Moreover, observations of Zipf's law have
  been obscured by word frequency distribution plots that fail to
  address correlated errors. Zipfian distributions are not unique to
  language. They appear in various aspects of human society, such as
  music and aspects of the internet, as well as in other biological and
  physical systems. Strikingly, power law distributions can arise from
  nearly nothing (\protect\hyperlink{ref-Piantadosi2014}{Piantadosi
  2014: 1121}).}
